### 2021-11-01 documents/sections/proto_carib.tex
2.1-2.3

\section{Cariban verbal person marking and marker extensions}
The archaic first person markers investigated in this paper are the result of incomplete person marker extensions, i.e., modifications to the ancestral prefix system that did not affect the whole lexicon.
To provide readers with the necessary background, we first introduce said ancestral system \pcref{sec:pc_person}, then the split-\gl{s} system which is very relevant for our topic \pcref{sec:split}, and finally the concept of person marker extensions \pcref{sec:extensions_intro}.

\subsection{Verbal person marking in \PC}
\label{sec:pc_person}
\PC is reconstructed by \textcite{gildea1998} as using a person paradigm called \setone in its independent verb forms, shown in \cref{tab:pcpers}.%
\footnote{We use standard IPA symbols in ourtranscription of Cariban languages, with the exception of coronal rhotics, which we simply represent with \ort{r}, rather than \ort{ɽ} for \wayana or \ort{ɾ̠} for \maqui etc.
In languages with strong morphophonological processes and/or subphonemic orthography we show the original transcription in an additional surface line when presented in an interlinearized glossed example.
We follow \textcite{gildea2018reconstructing} in using \ort{ə} for the proto-vowel reconstructed by \textcite{meira2005southern}, although it was likely more back \parencite{gildea2010story}.}
The choice of person marker in transitive verbs can be characterized as being conditioned by a basic person hierarchy \fbox{\gl{1}/\gl{2} > \gl{3}}.
The locuphoric markers had two forms, an \gl{a}-oriented one for direct (\gl{sap}>\gl{3}) scenarios and a \gl{p}-oriented one for inverse (\gl{3}>\gl{sap}) scenarios.
There was a single aliophoric marker \rc{n(i)-}, which only surfaced in nonlocal (\gl{3}>\gl{3}) scenarios, without morphologically expressed distinctions between different third person referents.
Local scenarios were expressed in a non-transparent manner, by using the \gl{1+2} prefix \rc{k-} in both cases.%
\footnote{The presence of a \gl{1+2} person value implies that of a \gl{1+3} value.
This is usually expressed with a free pronoun combined with third person morphology in Cariban languages, so it is not represented in this table.}

@@

\subsection{Defining features and origins of the split-\gl{s} system}
\label{sec:split}
As seen in the previous section, the split-\gl{s} defined two inflectional classes for intransitive verbs within the \setone system.
%\kalina paradigms illustrating the split for each person are given in \cref{tab:kalinaparadigms}.
%\gl{3}\gl{s_a_} differs from \gl{3}\gl{s_p_} by \begin{inlinelist}
%\item not showing the allomorph \obj{ni-} before consonants
%\item not triggering the umlaut of verb-initial \phon{\obj{o}}{\obj{e}}
%\end{inlinelist}, which is shown in \cref{tab:kalinaparadigms}.
%
%\begin{table}[h]
%	\centering
%	\caption{\kalina \gl{s_a_} and \gl{s_p_} person paradigms \parencites[178]{hoff1968carib}[430, 73]{courtz2008carib}[481]{meira2010origin}}
%	\label{tab:kalinaparadigms}
%	\begin{tabular}{@{}lll@{}}
%	\mytoprule
%Person & \gl{s_a_} \obj{opɨ} \qu{to come} & \gl{s_p_} \obj{[o/e]kanumɨ} \qu{to run}\\
%	\mymidrule
%\gl{1} & \obj{w-opɨ-} & \obj{j-ekanumɨ-}\\
%\gl{2} & \obj{m-opɨ-} & \obj{aj-ekanumɨ-}\\
%\gl{1+2} & \obj{kɨt-opɨ-} & \obj{k-okanumɨ-}\\
%\gl{3} & \obj{n-opɨ-} & \obj{n-ekajumɨ-}\\
%	\mybottomrule
%	\end{tabular}
%\end{table}
%
However, there were some other morphological criteria distinguishing \gl{s_a_} from \gl{s_p_} verbs in \PC:
\begin{inlinelist}
\item presence vs absence of the \gl{s_a_} marker \rc{w-}
\item absence vs presence of the second person prefix \rc{ə(j)-} in imperatives
\item presence vs absence of a derivational detransitivizing prefix
\end{inlinelist}.

@@

The distinction between \gl{s_a_} and \gl{s_p_} is also borne out in imperatives (\PC \rc{-kə}).
Here, \gl{s_p_} verbs took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while \gl{s_a_} verbs were unprefixed.
This is illustrated with reflexes in various modern languages in \cref{tab:imperatives}.
As in the case of the \gl{s_a_} marker \rc{w-} participles and nominalizations, some languages have lost the distinction between \gl{s_a_} and \gl{s_p_} verbs in imperatives, for example \panare.

@@

%\subsubsection{The detransitivizing prefix}
There is one further property uniting \gl{s_a_} verbs, which is not based on inflectional morphology.
As mentioned in \cref{sec:pc_person}, mismatches between the semantics of intransitive verbs and their \gl{a}- or \gl{p}-oriented inflectional morphology are common.
However, the Cariban split-\gl{s} system goes further than all other known such systems, in that its division of the verbal lexicon does not follow any discernible semantic criteria whatsoever.
\textcite{meira2000split} takes a sizable corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs.

Rather, the reason for a verb to take the \gl{a}- or \gl{p}-oriented prefix is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotation}
Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources) \parencite[201]{meira2000split}
\end{quotation}
\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
\begin{inlinelist}
	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
\end{inlinelist}.
He also notes that this leads to an inflectional split not based in meaning, but rather morphology:

\begin{quotation}
Apparently, the morphological behavior of the \gl{s_a_} verb class is an accidental consequence of the fact that detransitivization, as far back as we can reconstruct, entails all the morphology described […] as typical of \gl{s_a_} verbs. The alignment of person-marking prefixes appears not to be driven by any semantic forces in the language; it is as though they were being dragged by the evolution of the reflexive marker. \parencite[226]{meira2000split}
\end{quotation}

As for the form of this reflexive marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although they have since merged into a single morpheme in modern Cariban languages.
Modern reflexes of \detrz show a range of meanings, which can all be characterized as \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

@@

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides such modifications as for example losing the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
One could also imagine a scenario whereby verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs Meira sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial, thus explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system features markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the Parukotoan languages as an example.
We have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of ∅/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

@@

\hixka has preserved split-\gl{s} only in the second person prefixes, while \kaxui still shows variation in the first person and \gl{1+2} prefixes.
\waiwai, on the other hand, has lost the system entirely, which notably happened via distinct innovations at three different diachronic stages.
In this case, the loss of the inflectional classes also entailed the loss of the other morphological traces of the system.
That is, the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} was lost in \waiwai, as evidenced by the contrast between the \waiwai and \kaxui deverbal forms in \exref{wailoss.wai-158} and \exref{wailoss.kax-138}.
Similarly, the \gl{2}\gl{s_p_} prefix \obj{a-} was extended to imperatives of (former) \gl{s_a_} verbs \exref[wailoss.wai-112]{wailoss.hix-43}, but only C-initial ones \parencite[62]{waiwaihawkins1998}.

\pex<wailoss>
\a<wai-158> \waiwai \parencite[][98]{waiwaihawkins1998}\\
\begingl
\gla k-eɸɨrka-t͡ʃhe//
\glb \gl{1+2}-fall-\gl{advz}.after//
\glft \qu{after our fall}//
\endgl
\a<kax-138> \kaxui \parencite[][49]{schuring2018kaxuyana}\\
\begingl
\gla ku-w-ehurka-tpɨrɨ//
\glb \gl{1+2}-\gl{s_a_}-fall-\gl{nmlz}.\gl{pst}.\gl{pert}//
\glft \qu{our fall}//
\endgl
\a<wai-112> \waiwai \parencite[][177]{waiwaihawkins1998}\\
\begingl
\gla a-mo-ko//
\glb \gl{2}-come-\gl{imp}//
\glft \qu{come!}//
\endgl
\a<hix-43> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla m-omokɨ-no//
\glb \gl{2}\gl{s_a_}-come-\gl{imm}//
\glft \qu{You have come.}//
\endgl
\xe

@@

While different cases of loss of split-\gl{s} are discussed by \textcite[91--96]{gildea1998}, this paper focuses on a so far neglected aspect of these person marking extensions.
We argue that they are executed via lexical diffusion, characterized as a type of extension by \textcite[106--115]{harris1995historical}; this hypothesis is supported by three facts.

First of all, the variation in first person and \gl{1+2} prefixes described above for \kaxui is not completely free.
Rather, some verbs only allow for example first person \obj{k-}, but not \obj{j-}, while others can occur with both, which is the expected pattern in a lexical diffusion scenario.
In addition, this is speaker-dependent \perscommpar{Spike Gildea}, which is what one would expect from a change in progress.
%todo put in GOOD kaxui data
Second, while there is no detailed diachronic scenario for the switch of \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} in the Tiriyoan languages \pcref{sec:taranoan}, \textcite[111--112]{meira1998proto} argues that it must have happened gradually rather than instantaneously, and entailed both markers spreading at the same time.
Whether this gradual switch was along ordered lines or not, lexical diffusion must have played a role.
Finally, the process of lexical diffusion is seen most clearly where it was incomplete.
Not all person marker extensions spread through the entire lexicon, but rather stopped short of some verbs.
This is illustrated for \trio first person forms of \gl{s_a_} verbs in \exref{trio1}; regular \gl{s_a_} verbs take phonologically determined allomorphs \obj{s-} (\envr{}{\obj{e}}) and \obj{t-} (\envr{}{\obj{ə}}), but \obj{tən} \qu{to go} takes an \dbqu{irregular} or \dbqu{archaic} prefix \obj{wɨ-}; it was not affected by the spread of innovative \obj{t-}/\obj{s-} and therefore preserves the old prefix.

\ex<trio1> \trio \gl{1}\gl{s_a_} verbs \parencite[292, 294]{triomeira1999}\\
\begin{tabular}[t]{@{}ll@{}}
\obj{s-epɨ} & \qu{I bathed} \\
\obj{s-entapo} & \qu{I yawned}\\
\obj{t-əturu} & \qu{I talked}\\
\obj{t-əənɨkɨ} & \qu{I slept}\\
\obj{wɨ-tən} & \qu{I went}\\
\end{tabular}
\xe
%
This paper is primarily about these incomplete extensions, and the verbs that were not affected by them.

@@

That is, apparently ongoing changes like the situation discussed above for \kaxui will not be investigated in detail.
The same is true for extensions that affected the entire verbal lexicon, although we will briefly comment on them.
We have investigated the 19 person marker extensions we are aware of, but only 6 of them left a group of irregularly or archaically inflected verbs.
Interestingly, these verbs are always \gl{s_a_} verbs, and the irregular marker is always a first person one.
While there have been extensions for other person values as well, they never affect \gl{s_a_} verbs, only \gl{s_p_} ones, and they always affect the entire lexicon, at least based on the attested data.
Besides the completed extension of \rc{k-} to \gl{1}\gl{s_p_} and of \rc{tɨt-} to \gl{1+2}\gl{s_p_} in \PWai \pcref{fig:par_ext}, examples include the extension of \gl{1+2}\gl{s_a_} \obj{s(ɨ)-} (< \rc{kɨt-}) in \apalai \pcref{tab:apalai}, of \gl{2}\gl{s_a_} \obj{m(ɨ)-} in \panare \pcref{tab:panare},\footnote{The presence of the third person marker \obj{n-} for \gl{1+2} is due to the wholesale loss of that inflectional value.} or the extension of the entire \gl{s_a_} set in \waimiri \pcref{tab:waimiri}.

@@

The remainder of this paper is structured as follows:
In \cref{sec:extensions}, we discuss each of these six incomplete extensions individually, listing all attested unaffected verbs, and reconstructing proto-paradigms where necessary.
In \cref{sec:verbs}, we take a comparative look at the verbs that were not affected by these extensions and searches for possible motivations.
Finally, we discuss the findings and put them in a general context of language change and morphology in \cref{sec:discussion}.

>>>>

\section{Cariban verbal person marking and marker extensions}
The archaic first person markers investigated in this paper are the result of incomplete person marker extensions, i.e., modifications to the ancestral prefix system that did not affect the whole lexicon.
To provide readers with the necessary background, we first introduce said ancestral system \pcref{sec:pc_person}, then the peculiarities of the Cariban split-\gl{s} system \pcref{sec:split}, and finally the concept of person marker extensions \pcref{sec:extensions_intro}.

\subsection{Verbal person marking in \PC}
\label{sec:pc_person}
\PC is reconstructed by \textcite{gildea1998} as using a person paradigm called \setone in its independent verb forms, shown in \cref{tab:pcpers}.%
\footnote{We use standard IPA symbols in our transcription of Cariban languages, with the exception of coronal rhotics, which we simply represent with \ort{r}, rather than \ort{ɽ} for \wayana or \ort{ɾ̠} for \maqui etc.
In languages with strong morphophonological processes and/or subphonemic orthography we show the original transcription in an additional surface line when presented in an interlinearized glossed example.
We follow \textcite{gildea2018reconstructing} in using \ort{ə} for the proto-vowel reconstructed by \textcite{meira2005southern}, although it was likely more back \parencite{gildea2010story}.}
The choice of person marker in transitive verbs can be characterized as being conditioned by a basic person hierarchy \fbox{\gl{1}/\gl{2} > \gl{3}}.\footnote{Glossing abbreviations: \glossingAbbrevsComma.}
The locuphoric markers had two forms, an \gl{a}-oriented one for direct (\gl{sap}>\gl{3}) scenarios and a \gl{p}-oriented one for inverse (\gl{3}>\gl{sap}) scenarios.
There was a single aliophoric marker \rc{n(i)-}, which only surfaced in nonlocal (\gl{3}>\gl{3}) scenarios, without morphologically expressed distinctions between different third person referents.
Local scenarios were expressed in a non-transparent manner, by using the \gl{1+2} prefix \rc{k-} in both cases.%
\footnote{The presence of a \gl{1+2} person value implies that of a \gl{1+3} value.
This is usually expressed with a free pronoun combined with third person morphology in Cariban languages, so it is not represented in this table.}

@@

\subsection{Defining features and origins of the split-\gl{s} system}
\label{sec:split}
As seen in the previous section, the split-\gl{s} system defined two inflectional classes for intransitive verbs within the \setone system.
%\kalina paradigms illustrating the split for each person are given in \cref{tab:kalinaparadigms}.
%\gl{3}\gl{s_a_} differs from \gl{3}\gl{s_p_} by \begin{inlinelist}
%\item not showing the allomorph \obj{ni-} before consonants
%\item not triggering the umlaut of verb-initial \phon{\obj{o}}{\obj{e}}
%\end{inlinelist}, which is shown in \cref{tab:kalinaparadigms}.
%
%\begin{table}[h]
%	\centering
%	\caption{\kalina \gl{s_a_} and \gl{s_p_} person paradigms \parencites[178]{hoff1968carib}[430, 73]{courtz2008carib}[481]{meira2010origin}}
%	\label{tab:kalinaparadigms}
%	\begin{tabular}{@{}lll@{}}
%	\mytoprule
%Person & \gl{s_a_} \obj{opɨ} \qu{to come} & \gl{s_p_} \obj{[o/e]kanumɨ} \qu{to run}\\
%	\mymidrule
%\gl{1} & \obj{w-opɨ-} & \obj{j-ekanumɨ-}\\
%\gl{2} & \obj{m-opɨ-} & \obj{aj-ekanumɨ-}\\
%\gl{1+2} & \obj{kɨt-opɨ-} & \obj{k-okanumɨ-}\\
%\gl{3} & \obj{n-opɨ-} & \obj{n-ekajumɨ-}\\
%	\mybottomrule
%	\end{tabular}
%\end{table}
%
However, there were some other morphological criteria distinguishing \gl{s_a_} from \gl{s_p_} verbs in \PC:
\begin{inlinelist}
\item presence vs absence of the \gl{s_a_} marker \rc{w-}
\item absence vs presence of the second person prefix \rc{ə(j)-} in imperatives
\item presence vs absence of a derivational detransitivizing prefix
\end{inlinelist}.

@@

The distinction between \gl{s_a_} and \gl{s_p_} is also borne out in imperatives.
Here, \gl{s_p_} verbs took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while \gl{s_a_} verbs were unprefixed (both suffixed with \rc{-kə}).
This is illustrated with reflexes in various modern languages in \cref{tab:imperatives}.
As in the case of the \gl{s_a_} marker \rc{w-} participles and nominalizations, some languages have lost the distinction between \gl{s_a_} and \gl{s_p_} verbs in imperatives, for example \panare.

@@

%\subsubsection{The detransitivizing prefix}
There is one further property uniting most \gl{s_a_} verbs, not based on inflectional morphology.
As mentioned in \cref{sec:pc_person}, mismatches between the semantics of intransitive verbs and their \gl{a}- or \gl{p}-oriented inflectional morphology are common.
However, the Cariban split-\gl{s} system goes further than all other known such systems, in that its division of the verbal lexicon does not follow any discernible semantic criteria whatsoever.
\textcite{meira2000split} takes a sizable corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs.

Rather, the reason for a verb to take the \gl{a}- or \gl{p}-oriented prefix is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
\begin{inlinelist}
	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
\end{inlinelist}.
He also notes that this leads to an inflectional split not based in meaning, but rather morphology:

\begin{quotebox}{\parencite[226]{meira2000split}}
Apparently, the morphological behavior of the \gl{s_a_} verb class is an accidental consequence of the fact that detransitivization, as far back as we can reconstruct, entails all the morphology described […] as typical of \gl{s_a_} verbs. The alignment of person-marking prefixes appears not to be driven by any semantic forces in the language; it is as though they were being dragged by the evolution of the reflexive marker.
\end{quotebox}

As for the form of this marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although they have since merged into a single morpheme, apparently in all languages.
Modern reflexes of \detrz show a range of meanings, which can all be characterized as \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

@@

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides modifications like losing the \gl{s_a_} class marker \rc{w-} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
In principle, one could also imagine a scenario in which verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs Meira sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial, thus explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system involves markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the three Parukotoan languages as an example.
We have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of ∅/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

@@

\hixka has preserved split-\gl{s} only in the second person prefixes, while \kaxui still shows variation in the first person and \gl{1+2} prefixes.
\waiwai, on the other hand, has lost the system entirely, which notably happened via distinct innovations at three different diachronic stages.
In this case, the loss of the inflectional classes also entailed the loss of the other morphological traces of the split-\gl{s} system.
That is, the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} was lost in \waiwai, as evidenced by the contrast between the \waiwai and \kaxui deverbal forms in \exref{wailoss.wai-158} and \exref{wailoss.kax-138}.
Similarly, the \gl{2}\gl{s_p_} prefix \obj{a-} was extended to imperatives of (former) \gl{s_a_} verbs \exref[wailoss.wai-112]{wailoss.hix-43}, but only C-initial ones \parencite[62]{waiwaihawkins1998}.

\pex<wailoss>
\begin{multicols}{2}
\a<wai-158> \waiwai \parencite[][98]{waiwaihawkins1998}\\
\begingl
\gla k-eɸɨrka-t͡ʃhe//
\glb \gl{1+2}-fall-\gl{advz}.after//
\glft \qu{after our fall}//
\endgl
\a<kax-138> \kaxui \parencite[][49]{schuring2018kaxuyana}\\
\begingl
\gla ku-w-ehurka-tpɨrɨ//
\glb \gl{1+2}-\gl{s_a_}-fall-\gl{nmlz}.\gl{pst}.\gl{pert}//
\glft \qu{our fall}//
\endgl
\a<wai-112> \waiwai \parencite[][177]{waiwaihawkins1998}\\
\begingl
\gla a-mo-ko//
\glb \gl{2}-come-\gl{imp}//
\glft \qu{come!}//
\endgl
\a<hix-43> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla m-omokɨ-no//
\glb \gl{2}\gl{s_a_}-come-\gl{imm}//
\glft \qu{You have come.}//
\endgl
\end{multicols}
\xe

@@

While different cases of loss of split-\gl{s} are discussed by \textcite[91--96]{gildea1998}, this paper focuses on a so far neglected aspect of the person marking extensions resulting in this loss.
As a prerequisite for our story, we argue that they are executed via lexical diffusion, characterized as a type of extension by \textcite[106--115]{harris1995historical}; this hypothesis is supported by three facts.
First, the variation in first person and \gl{1+2} prefixes described above for \kaxui is not completely free.
Rather, some verbs only allow for example first person \obj{k-}, but not \obj{j-}, while others can occur with both, which is the expected pattern in a lexical diffusion scenario.
In addition, this is speaker-dependent \perscommpar{Spike Gildea}, which is what one would expect from a change in progress.
%todo put in GOOD kaxui data
Second, while there is no detailed diachronic scenario for the switch of \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} in the Tiriyoan languages \pcref{sec:taranoan}, \textcite[111--112]{meira1998proto} argues that it must have happened gradually rather than instantaneously, and entailed both markers spreading at the same time.
Whether this gradual switch was along ordered lines or not, lexical diffusion must have played a role.
Finally, the process of lexical diffusion is seen most clearly where it was incomplete.
Not all person marker extensions spread through the entire lexicon, but rather stopped short of some verbs.
This is illustrated for \trio first person forms of \gl{s_a_} verbs in \exref{trio1}; regular \gl{s_a_} verbs take phonologically determined allomorphs \obj{s-} (\envr{}{\obj{e}}) and \obj{t-} (\envr{}{\obj{ə}}), but \obj{tən} \qu{to go} takes an \dbqu{archaic} or \dbqu{irregular} prefix \obj{wɨ-}; it was not affected by the spread of innovative \obj{t-}/\obj{s-} and therefore preserves the old prefix.

\ex<trio1> \trio \gl{1}\gl{s_a_} verbs \parencite[292, 294]{triomeira1999}\\
\begin{tabular}[t]{@{}ll@{}}
\obj{s-epɨ} & \qu{I bathed} \\
\obj{s-entapo} & \qu{I yawned}\\
\obj{t-əturu} & \qu{I talked}\\
\obj{t-əənɨkɨ} & \qu{I slept}\\
\obj{wɨ-tən} & \qu{I went}\\
\end{tabular}
\xe
%
These incomplete extensions and the verbs that were not affected by them are the central topic of this paper.

@@

That is, apparently ongoing changes like the situation discussed above for \kaxui will not be investigated in detail.
The same is true for extensions that affected the entire verbal lexicon, although we will briefly comment on them.
We have investigated the 19 person marker extensions we are aware of, but only 6 of them left a group of irregularly or archaically inflected verbs.
Interestingly, these verbs are always \gl{s_a_} verbs, and the irregular marker is always a first person one.
While there have been extensions for other person values as well, they never affect \gl{s_a_} verbs, only \gl{s_p_} ones, and they always affect the entire lexicon, at least based on the attested data.
Illustrative examples for extensions affecting the entire lexicon are shown in \cref{tab:completed}: the extension of \gl{1+2}\gl{s_a_} \obj{s(ɨ)-} (< \rc{kɨt-}) in \apalai \pcref{tab:apalai}, of \gl{2}\gl{s_a_} \obj{m(ɨ)-} in \panare \pcref{tab:panare},\footnote{The presence of the third person marker \obj{n-} for \gl{1+2} is due to the wholesale loss of that inflectional value.} or the extension of the entire \gl{s_a_} set in \waimiri \pcref{tab:waimiri}.%, as well as the extension of \rc{k-} to \gl{1}\gl{s_p_} and of \rc{tɨt-} to \gl{1+2}\gl{s_p_} in \PWai \pcref{fig:par_ext}.

@@

The remainder of this paper is structured as follows:
In \cref{sec:extensions}, we discuss each of these six incomplete extensions individually, discussing all attested unaffected verbs, and reconstructing proto-paradigms where necessary.
In \cref{sec:verbs}, we take a comparative look at the verbs that were not affected by these extensions.
Finally, we search for factors motivating the resistance of these verbs, discuss the findings and put them in a general context of language change and morphology in \cref{sec:discussion}.
